---
title: "GLM Intro"
output: 
  html_document:
    toc: true
    toc_float: true
---

<style>
/* Table of Contents Styles */
#TOC {
  color: #333333;
  font-family: 'Arial', sans-serif; 
  font-size: 16px; 
  border-color: #333333;
}

/* Header Styles */
#header {
  color: #B22222; 
  background-color: #F9F9F9; 
  opacity: 0.9; 
  font-family: 'Arial', sans-serif;
  font-size: 24px;
  padding: 15px; 
}

/* Body Styles */
body {
  color: #000000; 
  font-family: 'Arial', sans-serif;
  background-color: #FFFFFF; 
  line-height: 1.6; 
}

/*  Text Styles */
pre {
  color: #4D4D4D; 
  background-color: #F0F8FF; 
  border: 1px solid #B0C4DE; 
  padding: 10px; 
}

/* Header Styles */
h1, h2, h3 {
  color: #003366;
  font-family: 'Arial', sans-serif;
}

/* Link Styles */
a {
  color: #003366; 
  text-decoration: none; 
}

a:hover {
  text-decoration: underline; 
}
</style>




## Probit


## Logistic Regression



```{r load packages, message=FALSE}
rm(list = ls()) 

library(dplyr)
library(data.table) 
library(haven)
library(margins) # handy functions to calculate marginal effects
library(lmtest) # helps cluster standard errors
library(sandwich) # variance-covariance estimators

options(scipen = 999) # makes numbers look nicer
```



```{r data load}

df <- setDT(read_dta("https://raw.githubusercontent.com/Michael-Morgan-Giles/regression_analysis/refs/heads/main/data/lakisha_aer.dta"))

df[,":=" (city_dummy = fifelse(city == "c", 1, 0),
      white_dummy = fifelse(race == "w", 1, 0),
      sex_dummy = fifelse(sex == "m", 1, 0),
      hl_dummy = fifelse(h == 1, "high", "low")
        )]

cols <- c("white_dummy",
     "h",
     'city_dummy',
     'sex_dummy',
     'honors',
     'volunteer',
     'military', 
     'email',
     'empholes',
     'workinschool', 
     'computerskills', 
     'specialskills',
     'yearsexp')

formula <- as.formula(paste0("call ~ ", paste0(cols, sep = " ", collapse = "+ ")))

print(formula)

```




```{r recreate key stats}
df[, .(count = .N,
       mean_callback_rate = round(mean(call),4),
       sd_callback_rate = round(sd(call),4)),
       by = .(race)]

```

```{r recreate key stats with quality}
dcast(df[, .(mean_callback_rate = round(mean(call), 4)),
       by = .(race, hl_dummy)],
      race ~ hl_dummy
) %>% 
  mutate(delta = high - low)

```

```{r t test main}

t_res <- data.frame()

for (c in cols[-1]){
  t_res_temp <- t.test(df[race == 'b', ..c],
                       df[race == 'w', ..c])
  
  t_res <- rbind(t_res,
                 data.frame(covariate = c, 
                            b = round(t_res_temp$estimate[1],4),
                            w = round(t_res_temp$estimate[2],4),
                            # delta = ,
                            t_stat = round(t_res_temp$statistic,4),
                            p_value = round(t_res_temp$p.value,4)))
}

#kableExtra::kable(t_res)
t_res
```



```{r t test resume quality}

t_res_quality <- data.frame()

for (c in cols[-2]){
  t_res_temp <- t.test(df[hl_dummy == 'high', ..c],
                       df[hl_dummy == 'low', ..c])
  
  t_res_quality <- rbind(t_res_quality,
                         data.frame(covariate = c,
                                    high = round(t_res_temp$estimate[1],4),
                                    low = round(t_res_temp$estimate[2],4),
                                    # delta = ,
                                    t_stat = round(t_res_temp$statistic,4),
                                    p_value = round(t_res_temp$p.value,4)))
}

#kableExtra::kable(t_res_quality)
t_res_quality
```

```{r factor}
# reformat to factor before fitting to model (helps with magrinal effect ccalcs)
df[,":="(white_dummy = as.factor(white_dummy),
                                  h = as.factor(h),
                                  city_dummy = as.factor(city_dummy),
                                  sex_dummy = as.factor(sex_dummy),
                                  honors = as.factor(honors),
                                  volunteer = as.factor(volunteer),
                                  military = as.factor(military),
                                  email = as.factor(email),
                                  empholes = as.factor(empholes),
                                  workinschool = as.factor(workinschool),
                                  computerskills = as.factor(computerskills),
                                  specialskills = as.factor(specialskills))]

```


```{r probit}
probit_res <- glm(formula = call ~ white_dummy + h + city_dummy + sex_dummy + honors + 
                    volunteer + military + email + empholes + workinschool + 
                    computerskills + specialskills + yearsexp, 
                  family = binomial(link = "probit"), 
                  data = df)

coeftest(probit_res, vcov. = vcovCL(probit_res, cluster = ~adid))

ame_probit <- margins(probit_res, 
                      #vce = c("bootstrap"), 
                      vcov = vcovCL(probit_res, cluster = ~adid))

summary(ame_probit)
```


```{r logit}
logit_res <- glm(formula = call ~ white_dummy + h + city_dummy + sex_dummy + honors + 
                    volunteer + military + email + empholes + workinschool + 
                    computerskills + specialskills + yearsexp, 
                  family = binomial(link = "logit"), 
                  data = df)

# cluster at ad id level
coeftest(logit_res, vcov. = vcovCL(logit_res, cluster = ~adid))

ame_logit <- margins(logit_res, 
                     # vce = c("bootstrap"), 
                      vcov = vcovCL(logit_res, cluster = ~adid))

summary(ame_logit)
```






