---
title: "Time Series Introduction"
output: 
  html_document:
    toc: true
    toc_float: true
---

<style>
/* Table of Contents Styles */
#TOC {
  color: #333333;
  font-family: 'Arial', sans-serif; 
  font-size: 16px; 
  border-color: #333333;
}

/* Header Styles */
#header {
  color: #B22222; 
  background-color: #F9F9F9; 
  opacity: 0.9; 
  font-family: 'Arial', sans-serif;
  font-size: 24px;
  padding: 15px; 
}

/* Body Styles */
body {
  color: #000000; 
  font-family: 'Arial', sans-serif;
  background-color: #FFFFFF; 
  line-height: 1.6; 
}

/*  Text Styles */
pre {
  color: #4D4D4D; 
  background-color: #F0F8FF; 
  border: 1px solid #B0C4DE; 
  padding: 10px; 
}

/* Header Styles */
h1, h2, h3 {
  color: #003366;
  font-family: 'Arial', sans-serif;
}

/* Link Styles */
a {
  color: #003366; 
  text-decoration: none; 
}

a:hover {
  text-decoration: underline; 
}
</style>










```{r load packages, message=FALSE}
rm(list = ls()) 

library(dplyr)
library(data.table) 
library(ggplot2)
library(fastDummies)

options(scipen = 999) # makes numbers look nicer
```

Great now lets load in some data for us to play around with. 

This dataset is a very simple collection of house sales. We will try to model some prices to see which features of houses are most important in determining prices *(as well as how helpful they are at predicting house prices)*.

```{r data load}

df <- setDT(read.csv("https://raw.githubusercontent.com/Michael-Morgan-Giles/python-causality-handbook/refs/heads/master/causal-inference-for-the-brave-and-true/data/smoking.csv"))

# filter for california
df <-  df[california == "True"]

# add binary variable for excise tax intro
df[, excise_tax_intro :=  fifelse(after_treatment == TRUE,1,0)]

df <-  df[, .(year, 
              cigsale,
              retprice,
              excise_tax_intro, 
              ln_cigsale = log(cigsale),
              ln_retprice = log(retprice))]

```





```{r cigsale plt}
df %>% 
  ggplot(aes(x = year, y = cigsale))+
  geom_line(color = "blue")+
  geom_vline(xintercept = 1989)+
  theme_classic()

```

```{r simple static model}
simple_static_mod <- lm(cigsale ~ excise_tax_intro + ln_retprice, data = df) 

summary(simple_static_mod)

df[, simple_mod_pred := predict(simple_static_mod)]

df %>% 
  ggplot(aes(x = year))+
  geom_line(aes(y = cigsale), color = "blue")+
  geom_line(aes(y = simple_mod_pred), color = "orange")+
  geom_vline(xintercept = 1989)+
  theme_classic()


df %>% 
  ggplot(aes(x = cigsale, y = simple_mod_pred))+
  geom_point(color = "blue")+
  theme_classic()
```

```{r complex model}
# add dummy varaibles for years
df <- dummy_cols(df, select_columns = "year")

complex_static_mod <- lm(as.formula(paste0("cigsale ~ excise_tax_intro + ln_retprice + ",  
                           paste0("year_",as.character(levels(as.factor(df[year > 1987, year]))), sep = " ", collapse = "+"))), 
                         data = df) 

summary(complex_static_mod)

df[, complex_mod_pred := predict(complex_static_mod)]

df %>% 
  ggplot(aes(x = year))+
  geom_line(aes(y = cigsale), color = "blue")+
  geom_line(aes(y = simple_mod_pred), color = "darkgreen")+
  geom_line(aes(y = complex_mod_pred), color = "orange")+
  geom_vline(xintercept = 1989)+
  theme_classic()


df %>% 
  ggplot(aes(x = cigsale, y = complex_mod_pred))+
  geom_point(color = "blue")+
  theme_classic()

```

```{r data load new}

df_new <- setDT(read.csv("https://raw.githubusercontent.com/Michael-Morgan-Giles/python-causality-handbook/refs/heads/master/causal-inference-for-the-brave-and-true/data/smoking.csv"))

# filter for california
df_new <-  df_new[state == 36]

# add binary variable for excise tax intro
df_new[, excise_tax_intro :=  fifelse(after_treatment == TRUE,1,0)]

df_new <-  df_new[, .(year, 
              cigsale,
              retprice,
              excise_tax_intro, 
              ln_cigsale = log(cigsale),
              ln_retprice = log(retprice))]

df_new <- dummy_cols(df_new, select_columns = "year")

df_new[, new_state_mod_pred := predict(complex_static_mod,
                                       df_new[,.(excise_tax_intro,
                                                 ln_retprice,
                                                 year_1988,
                                                 year_1989,
                                                 year_1990,
                                                 year_1991,
                                                 year_1992,
                                                 year_1993,
                                                 year_1994,
                                                 year_1995,
                                                 year_1996,
                                                 year_1997,
                                                 year_1998,
                                                 year_1999,
                                                 year_2000)])]

df_new %>% 
  ggplot(aes(x = year))+
  geom_line(aes(y = cigsale), color = "blue")+
  geom_line(aes(y = new_state_mod_pred), color = "orange")+
  geom_vline(xintercept = 1989)+
  theme_classic()

```



