---
title: "Simple Exploration of Optimisation"
output: 
  html_document:
    toc: true
    toc_float: true
---

<style>
/* Table of Contents Styles */
#TOC {
  color: #333333;
  font-family: 'Arial', sans-serif; 
  font-size: 16px; 
  border-color: #333333;
}

/* Header Styles */
#header {
  color: #B22222; 
  background-color: #F9F9F9; 
  opacity: 0.9; 
  font-family: 'Arial', sans-serif;
  font-size: 24px;
  padding: 15px; 
}

/* Body Styles */
body {
  color: #000000; 
  font-family: 'Arial', sans-serif;
  background-color: #FFFFFF; 
  line-height: 1.6; 
}

/*  Text Styles */
pre {
  color: #4D4D4D; 
  background-color: #F0F8FF; 
  border: 1px solid #B0C4DE; 
  padding: 10px; 
}

/* Header Styles */
h1, h2, h3 {
  color: #003366;
  font-family: 'Arial', sans-serif;
}

/* Link Styles */
a {
  color: #003366; 
  text-decoration: none; 
}

a:hover {
  text-decoration: underline; 
}
</style>


# Simple Exploration of Optimisation
## Maximum Likelihood Estimator

So far throughout these notebooks we've been using the analytical solution to estimate $\beta$, via OLS: 

$$ \hat\beta  = (XX')^{-1}XY $$

This is obviously fine and can be an efficient way to estimate $\beta$. However, OLS can be reframed as one unique solution to another slightly more general way of estimating parameters - **Maximum Liklihood Estimation (MLE)**. 

MLE is basically an optimisation problem that chooses parameters, $\beta$, that are the maximum a models likelihood function. Likelihood functions describe the probability that observed data, $Y$, is generated by a set of observed data, $X$, and some parameters $\beta$, i.e. $\Pr(Y|X,\beta)$.

The parameters are choosen based on the solution to the likelihood function that will maximize the porbability of the model producing data that is the same as the observed data points. Whilst not the only way to fit parameters of models, it is a highly popular option manily because of it's abilty to not require distrbutional assumptions for model errors. Really handily for our purposes we can actually reframe all our previous OLS estimates into MLE. 

So let's explore what's going on under the hood!

```{r load packages, message=FALSE}
rm(list = ls()) 

library(dplyr)
library(data.table) 
library(ggplot2)
library(plotly)

options(scipen = 999) # makes numbers look nicer
```


Note that all of the null wage figures are for those who are not in the labour force. Therefore, I am just going to assume those people have no wage and update those nulls to zero.

```{r data load}

df <- setDT(read.csv("https://raw.githubusercontent.com/Michael-Morgan-Giles/regression_analysis/refs/heads/main/data/mroz9.csv"))

df[is.na(lnwage), lnwage := 0]

df[, sum(is.na(lnwage))]

formula <- as.formula("inlf ~ nwifeinc + educ + exper + expersq + age + kidslt6 + lnwage")
```

### Linear Regression as MLE

As mentioned, we can use MLE to choose a vector $ \beta$ that maximises the probability of the observed data matching the output of the model.

So, say we have our typical linear model $Y = \beta X + \varepsilon$, and we want to maximise the probabiliy that $\beta$ produces $ Y $ given $X$, i.e.

$$ \Pr(Y | X, \beta) $$ 

Now this can become quite a difficult function to optimise *(finding the derivatives analytically is not trivial)*, so we tend to optimise the log of this function i.e.

$$ \ln  \Pr(Y | X, \beta) $$

This makes the calculus much easier to manage *(it changes products to sums)*. So now the MLE estimator is basically just:

$$ \beta_{MLE} \rightarrow \max_{\beta} \ln  \Pr(Y | X, \beta) $$ 

Now, even more handily for us, any maximization problem can be turned into a minimisation problem by maximising the negative of the problem. For us, this means we can use standard minimisation functions to solve the following problem:

$$ \beta_{MLE} \rightarrow \max_{\beta} \ln  \Pr(Y | X, \beta) = \min_{\beta} - \ln  \Pr(Y | X, \beta) $$

I won't get too much into the direct derivation of the log-likelihood funciton for a linear model *(for more check this article out: https://statproofbook.github.io/P/slr-mle.html - note i've simplified it a bit)*. For this notebook, we can use the following function to esitmate $\beta$ *(and the variance parameter $\sigma$)*:

$$ 
nll(\beta, \sigma) =  \frac{n}{2}\ln(2\pi\sigma^2) + \frac{1}{2\sigma^2}\sum_{i=1}^n (Y - X\beta)^2 
$$   


We can use the `optim()` functions from the base `stats` package to fit the parameters. Let's code this up and see if there are any differences to either approach. Firstly, lets estimate the model using OLS:

```{r baseline ols}

round(summary(lm(formula, data = df))$coefficients,4)
```

Okay great, we now have a baseline of $\beta$ to compare to our minimisation outputs below.

Let's compare this to minimising the log-likelihood function and see if we get the same vector of parameters!

```{r optim}
negative_ll <- function(params, x, y, n){
  b <- params[-length(params)] 
  sd <- params[length(params)] 
  
  r = y - x%*%b
  r2 <- t(r) %*% r
  
  return((n / 2) * log(2 * pi * sd^2) + (r2 / (2 * sd^2)))
}

```

Now before we get to the minimisation problem itself, we need to set some things up.

Firstly, lets define our matrix $X$. Secondly, we need to determine both the number of parameters we want to estimate for $\beta$, but also we need to set up some initial guesses for them. 

This tells the optimisation problem where to start the minimisation algorithm from. Note that we add an extra element to $\beta$ so we can calculate an estimate for $\sigma$ as the this last element. We will start the guess at 1 because a standard deviation of zero is extremely unlikely. 

The last little thing we need to set up for this linear model is to ensure that $\sigma$ cannot be negative. This involves using the `lower` argument in the `optim()` function. This sets the constraints on where the minimisation algorithm will search for each element in $\beta$. We will ensure that $\sigma > 0$ *(set to a really low number but not zero directly)*.

Now that we have the data, we know the parameters we want to estimate, the constraints on those parameters, lets minimize the negative Log-likelihood function!

```{r initial parameters}

X_matrix <- as.matrix(df[, .(const = 1, nwifeinc, educ, exper, expersq, age, kidslt6, lnwage)])
y_matrix <- df$inlf
n_params <- ncol(X_matrix)

# optimisation problem
lin_res <- optim(par = c(rep(0, n_params), 1),
                 fn = negative_ll,
                 x = X_matrix,
                 y = y_matrix,
                 n = nrow(X_matrix),
                 method  = "L-BFGS-B",
                 lower = c(rep(-Inf, n_params), 1e-20),
                 upper = c(rep(Inf, n_params + 1)),
                 control = list(maxit = 20000, factr = 1e-10),
                 hessian = FALSE)

print(lin_res)

```

You can see in the print out above some interesting information about the optimisation problem including the function value `vale`, and some other helpful information you can return. 

But perhaps of the most interest is the parameter outputs `par`:

```{r print par}
knitr::kable(lin_res$par)
```

Yay they're identical!  You will note there is an extra parameter in the `optim()` output, `par`. This is the estimated $\hat \sigma$ parameter estimate. 

Interestingly, it is every so slightly different than the the `lm()` output, however it is arbitrarily close to the to each other.

```{r sigma dif}
print(summary(lm(formula, data = df))$sigma)
```

The basic reason for this is that the analytical OLS solution assumes an extra degree of freedom than the MLE estimate. This means $\sigma$ will always be slightly underestimated in the MLE solutions, but that's a bit beyond what we need to get into here. 

For now, we can see how the different estimates for $\beta$ converge to the same parameters, despite being estimated in two different ways! 

Let's use MLE in a slightly more complex and different case - a classification / binary choice problem!

### A (very) brief intro to classification problems

So we've seen we can estimate parameters in a linear regression using MLE. But as mentioned, MLE is also a helpful estimator in other scenarios - perhaps most noteably in classification problems, especially for Logistic Regressions. 

I will not go into depth with logistic regressions *(that is for the next notebook)*, however for our purposes a logistic regression esimtates the probabilty of a binary outcome variable $Y$, given a set of features $X$ and a vector of parameters $\beta$. 

This should sound very familiar. Indeed they're relatively similar to linear regressions except are used for classification instead of regression problems i.e. for discrete outcomes such that $Y \in 0,1$. They basically transform linear predictions into a probability score that is bounded by 0 or 1 using a transformation known as a sigmoid. This will all become more clear in the next notebook. 

For now, we can use the negative log-likelihood functions for a logistic regression to comapre against the typical stats model estimates.

The negative likelihood function of a logistic model is: 

$$nll(\beta) =  - \sum -\ln 1 + e^{X\beta} + yX\beta$$

For derivation see: https://www.statlect.com/fundamentals-of-statistics/logistic-model-maximum-likelihood. 

Okay great, we've for the negative log-liklihood function. Let's code this up! 

But again, lets first use the same data from the problem above, but use `glm` to get a baseline estimate for $\beta$: 

```{r baseline glm res}

round(summary(glm(formula = formula, family = binomial(), data = df))$coefficients,4)

```

Cool we have the parameters for $\beta$. So similarly to the linear model, lets first define the negative log-likelihood function.

In the same way we can put in zeros for the initial guesses of $\beta$ plus an extra zero for the intercept. 

But note we are not estimating a $\sigma$ in this problem, and we don't have any constraints set on the coefficients *(like in the OLS example)*. 

So we can just get straight into the optimisation problem!

```{r minimisation of log reg}
logistic_nll <- function(b, y, x){
  return(-sum(-log(1 + exp(x %*% b)) + y * x %*% b))
}

log_res <- optim(par = matrix(c(rep(0, ncol(X_matrix))), ncol = 1),
                 fn = logistic_nll,
                 y = y_matrix,
                 x = X_matrix,
                 control = list(maxit = 10000,
                                reltol = 1e-10),
                 method = "BFGS")

print(log_res)

knitr::kable(log_res$par)

```

We see that the coefficients are both approaches the same numbers, yay! *(ignore the slight differences)*

Now, this is fun and all, but what's quite exciting is that MLE can be used to estimate all sorts of parameters. For example, it can fit a distribution such as the gamma, beta or log-normal distributions. 

It can fit most systems of equations you make. 

It's all quite flexible, and now with a rudimentary understanding of what we are trying to achieve with MLE, you can try apply it in other scenarios!


# Numerical Optimisation

Now that we've explored MLE, there are other ways to optimise a models parameters. One very common in machine learning / statistics more generally are numerical methods. 

Very basically, numerical methods use the ability of computers to quickly transform large data structures using algorithms. In fact, we've actually been using this to minimise the log-likelihood functions already *(note the `method` argument we've been setting this whole time)*. 

However, instead of optimising the log-likelihood function, we can also use numerical methods to directly minimise a "cost" or "loss" function. 

As a simple example let's optimise a linear model by minimise the sum of squared residuals *(hopefully that sounds familiar!)*:

$$ SSR = (y - \beta X)'(y - \beta X) = \|y - \beta X\|_2$$

Using the same dataset above, we can find the coefficients by minimising the above cost / loss function: 


```{r ssr}
ssr_loss <- function(b,y,x){
  return(sum((y - x%*%b)^2))
}

ssr_res <- optim(par = c(rep(0, ncol(X_matrix))),
                 fn = ssr_loss,
                 y = y_matrix,
                 x = X_matrix,
                 method = "BFGS")

knitr::kable(ssr_res$par)

```

As you can see the vector is the same as the one calculated above. This can be quite handy if you don't know the log-likeihood function. This is because you can generally identify some loss function to minimise such as mean squared error (MSE) quite easily. 

The really cool thing about numerical optimisation is that you can fit really complicated functions quite easily *(especially those that can be difficult to solve analytically)*.

For example pick any complicated looking function and see if we can optimise it! Say maybe this interesting looking function: 

$$ f(x,y) = (x^2 + y - 11)^2 + (x + y^2 - 7)^2 $$

```{r himmel}
himmel <- function(params){
  x <- params[1]
  y <- params[2]
  
  return( (x^2 + y - 11)^2 + (x + y^2 - 7)^2 )
}

himmel_res <- optim(fn = himmel,
      par = c(4,1),
      method = "BFGS")

print(himmel_res)

knitr::kable(himmel_res$par)
```



```{r 3d plot}
# helper func for plot
himmel_plot <- function(x,y){
  return( (x^2 + y - 11)^2 + (x + y^2 - 7)^2 )
}

x <- seq(-5, 5, length.out = 100)
y <- seq(-5, 5, length.out = 100)

grid <- expand.grid(x = x, y = x)

z <- matrix(himmel_plot(grid$x, grid$y), nrow = 100, ncol = 100)

plot_ly(x = x,
        y = y,
        z = ~z,
        type = "surface") %>% 
  add_trace(
  type = "scatter3d",
  mode = "markers",
  x = ~himmel_res$par[1],
  y = ~himmel_res$par[2],
  z = ~himmel(himmel_res$par),
  marker = list(
    size = 6,     
    color = "red",
    opacity = 0.8
  ),
  name = "Minimum point for function"
)
```

As you can see this actually turns out to look quite complex when you plot the surface of the function. This is actually a Himmelblau function. It is often used in mathematical optimisation because it is multimodal i.e. has many local minima. 

The one we found is just one of four. Maybe you can play around and see if you can find the other ones *(hint, you might need to update the initial guesses)*.

It is worth noting that despite being able to fit complex functions quite easily, it can come at quite a high computational cost. But at small scales like this it's quite insignificant *(this is really only a problem when working with really big sets of high dimensional data - but is always worth considering before pressing run)*.

Hopefully this brief introduction into optimisation has helped illuminate what is going on under the hood when someone says they have "fitted", "estimated" or "trained" a model, and not they've just spit out numbers magically.
