---
title: "Simple Exploration of Optimisation"
output: 
  html_document:
    toc: true
    toc_float: true
---

<style>
/* Table of Contents Styles */
#TOC {
  color: #333333;
  font-family: 'Arial', sans-serif; 
  font-size: 16px; 
  border-color: #333333;
}

/* Header Styles */
#header {
  color: #B22222; 
  background-color: #F9F9F9; 
  opacity: 0.9; 
  font-family: 'Arial', sans-serif;
  font-size: 24px;
  padding: 15px; 
}

/* Body Styles */
body {
  color: #000000; 
  font-family: 'Arial', sans-serif;
  background-color: #FFFFFF; 
  line-height: 1.6; 
}

/*  Text Styles */
pre {
  color: #4D4D4D; 
  background-color: #F0F8FF; 
  border: 1px solid #B0C4DE; 
  padding: 10px; 
}

/* Header Styles */
h1, h2, h3 {
  color: #003366;
  font-family: 'Arial', sans-serif;
}

/* Link Styles */
a {
  color: #003366; 
  text-decoration: none; 
}

a:hover {
  text-decoration: underline; 
}
</style>










```{r load packages, message=FALSE}
rm(list = ls()) 

library(dplyr)
library(data.table) 
library(ggplot2)
library(plotly)

options(scipen = 999) # makes numbers look nicer
```



```{r data load}

df <- setDT(read.csv("https://raw.githubusercontent.com/Michael-Morgan-Giles/regression_analysis/refs/heads/main/data/mroz9.csv"))

df[is.na(lnwage), lnwage := 0]

df[, sum(is.na(lnwage))]

formula <- as.formula("inlf ~ nwifeinc + educ + exper + expersq + age + kidslt6 + lnwage")
```

https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/optim


```{r baseline ols and optim}

round(summary(lm(formula, data = df))$coefficients,4)

negative_ll <- function(params, x, y, n){
  b <- params[-length(params)] 
  sd <- params[length(params)] 
  
  r = y - x%*%b
  r2 <- t(r) %*% r
  
  return((n / 2) * log(2 * pi * sd^2) + (r2 / (2 * sd^2)))
}

X_matrix <- as.matrix(df[, .(const = 1, nwifeinc, educ, exper, expersq, age, kidslt6, lnwage)])
y_matrix <- df$inlf
n_params <- ncol(X_matrix)


optim(par = c(rep(0, n_params), 1),
      fn = negative_ll,
      x = X_matrix,
      y = y_matrix,
      n = nrow(X_matrix),
      method  = "L-BFGS-B",
      lower = c(rep(-Inf, n_params), 1e-20),
      upper = c(rep(Inf, n_params + 1)),
      control = list(maxit = 10000, factr = 1e-10),
      hessian = TRUE)

```


```{r logistic regression}

summary(glm(formula = formula, family = binomial(), data = df))$coefficients

logistic_nll <- function(b, y, x){
  return(-sum(-log(1 + exp(x %*% b)) + y * x %*% b))
}

optim(par = matrix(c(rep(0, ncol(X_matrix))), ncol = 1),
      fn = logistic_nll,
      y = y_matrix,
      x = X_matrix,
      control = list(maxit = 10000, 
                     reltol = 1e-10),
      method = "BFGS")


logistic_nll(b = matrix(c(rep(0, ncol(X_matrix))), ncol = 1), y = y_matrix, x = X_matrix)

```




# Numerical Optimisation

```{r ssr}
ssr_loss <- function(b,y,x){
  return(sum((y - x%*%b)^2))
}

optim(par = c(rep(0, ncol(X_matrix))),
      fn = ssr_loss,
      y = y_matrix,
      x = X_matrix,
      method = "BFGS")

```



```{r himmel}
himmel <- function(params){
  x <- params[1]
  y <- params[2]
  
  return( (x^2 + y - 11)^2 + (x + y^2 - 7)^2 )
}

himmel_res <- optim(fn = himmel,
      par = c(0,0),
      method = "BFGS")

himmel_res$par
```



```{r 3d plot}
himmel_plot <- function(x,y){
  return( (x^2 + y - 11)^2 + (x + y^2 - 7)^2 )
}

x <- seq(-5, 5, length.out = 100)
y <- seq(-5, 5, length.out = 100)

grid <- expand.grid(x = x,
            y = x)

z <- matrix(himmel_plot(grid$x, grid$y), nrow = 100, ncol = 100)

plot_ly(x = x,
        y = y,
        z = ~z,
        type = "surface") %>% 
  add_trace(
  type = "scatter3d",
  mode = "markers",
  x = ~himmel_res$par[1],
  y = ~himmel_res$par[2],
  z = ~himmel(himmel_res$par),
  marker = list(
    size = 6,     
    color = "red",
    opacity = 0.8
  ),
  name = "Minimum point for function"
)
```


