---
title: "Intro to Regression"
output:
  html_document: default
---

# Introduction to Regression

What is a regression? At its most basic, it is a method of fitting some data, $X$ that can explain / predict some other data $Y$. Note that regressions tend to be for continuous data (discrete data is another set of methods that tend to be known as classification). 

So, we want to learn how some set of features in $X$ relate to an outcome $Y$. Simple enough to start.

Let’s explore how we can estimate this relationship that tends to be represented by the vector $\beta$.

I will be exploring linear models in matrix notation because I tend to find it much easier and less tedious. This also allows us to then focus in it’s interpretation and implementation.

However, in order for us to be on the same page when using matrix notation, we need to get the basic tedious background of matrix notation out of the way! 

So let's do a quick refresh on matrices.

## A (very) brief introduction to linear models in matrix form

### 1. Basic Definitions
Firstly, what is a matrix? 

Well, a matrix is a rectangular array of numbers with dimensions $r \times c$.
*   A matrix is **square** if $r = c$.
*   A matrix is **symmetrical** if $a_{ij} = a_{ji} \quad \forall i,j$.

A general matrix $X$ is represented as:
$$
X = \begin{bmatrix}
a_{11} & a_{12} & \cdots & a_{1c} \\
a_{21} & a_{22} & \cdots & a_{2c} \\
a_{31} & a_{32} & \cdots & a_{3c} \\
\vdots & \vdots & \ddots & \vdots \\
a_{r1} & a_{r2} & \cdots & a_{rc}
\end{bmatrix}
$$

### The Main Diagonal
The main (or principal) diagonal of a square matrix (where $r=c$) is:
$$ \text{diag}(A) = (a_{11}, a_{22}, a_{33}, \dots, a_{cc}) $$

### Diagonal Matrix
A square matrix is **diagonal** if $a_{ij} = 0 \quad \forall i \neq j$:

$$
X = \begin{bmatrix}
a_{11} & 0 & 0 & \cdots & 0 \\
0 & a_{22} & 0 & \cdots & 0 \\
0 & 0 & a_{33} & \cdots & 0 \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
0 & 0 & 0 & \cdots & a_{cc}
\end{bmatrix}
$$

### Identity Matrix
The above matrix is an **identity matrix**, $I_n$, if the $n \times n$ matrix diagonal has $a_{ii} = 1, \quad \forall i=1,\dots,n$.

## 2. Matrix Operations

### Addition and Subtraction
To add or subtract matrices, the dimensions of each matrix must be identical.
*   $A = B$ if $a_{ij} = b_{ij} \quad \forall i,j$
*   $C = A + B = [a_{ij} + b_{ij}]$

$$
X = \begin{bmatrix}
a_{11}+b_{11} & a_{12}+b_{12} & \cdots & a_{1c}+b_{1c} \\
a_{21}+b_{21} & a_{22}+b_{22} & \cdots & a_{2c}+b_{2c} \\
a_{31}+b_{31} & a_{32}+b_{32} & \cdots & a_{3c}+b_{3c} \\
\vdots & \vdots & \ddots & \vdots \\
a_{r1}+b_{r1} & a_{r2}+b_{r2} & \cdots & a_{rc}+b_{rc}
\end{bmatrix}
$$

### Transpose
To transpose a matrix is to flip the rows and columns. $A'$ (or $A^T$) denotes the transpose.
*   By definition: $A = (A')'$
*   If $A$ is $(r \times c)$, then $A'$ is $(c \times r)$.
*   $A$ is **symmetric** if $A = A'$.

### Vectors
*   A **row vector** is a $1 \times c$ matrix.
*   Its transpose is a **column vector** ($r \times 1$):
$$
x' = \begin{pmatrix} x_1 \\ x_2 \\ \vdots \\ x_r \end{pmatrix}, \quad x = (x_1, x_2, \dots, x_c)
$$

### Inner Product
The inner product of a $1 \times k$ row vector and a $k \times 1$ column vector is defined as:
$$
x'\beta = (x_1, x_2, \dots, x_k) \begin{pmatrix} \beta_1 \\ \beta_2 \\ \vdots \\ \beta_k \end{pmatrix} = \sum_{j=1}^k x_j \beta_j
$$

### Matrix Decomposition
Matrices can be viewed as rows of $k$ vectors or columns of $n$ vectors:
$$
X = \begin{bmatrix} a_{11} & \cdots & a_{1c} \\ \vdots & \ddots & \vdots \\ a_{r1} & \cdots & a_{rc} \end{bmatrix} = \begin{bmatrix} \mathbf{a}_1 & \mathbf{a}_2 & \cdots & \mathbf{a}_c \end{bmatrix}
$$
$$
X = \begin{bmatrix} a_{11} & \cdots & a_{1c} \\ \vdots & \ddots & \vdots \\ a_{r1} & \cdots & a_{rc} \end{bmatrix} = \begin{pmatrix} \mathbf{a}^1 \\ \mathbf{a}^2 \\ \vdots \\ \mathbf{a}^r \end{pmatrix}
$$

### Multiplication
To multiply a matrix, the number of rows in $A_{(n \times p)}$ must equal the number of columns in $B_{(p \times q)}$. The result is a matrix $C_{(n \times q)}$:
$$
AB = (a_{n1}, \dots, a_{np}) \begin{pmatrix} b_{1q} \\ \vdots \\ b_{pq} \end{pmatrix} = \sum_{j=1}^p a_{nj} b_{jq} = c_{nq}
$$

#### Vector Multiplication Visualization
When multiplying a $u_{(n \times 1)}$ vector, the result depends on the order:

1.  **Scalar Result ($u'u$):**
    $$
    u'u = (u_1, \dots, u_n) \begin{pmatrix} u_1 \\ \vdots \\ u_n \end{pmatrix} = \sum_{i=1}^n u_i^2 = \text{scalar}
    $$
    *Note: When $u$ is a regression error term, $u'u$ is the Sum of Squared Errors (often related to MSE).*

2.  **Matrix Result ($uu'$):**
    $$
    uu' = \begin{pmatrix} u_1 \\ \vdots \\ u_n \end{pmatrix} (u_1, \dots, u_n) = \begin{bmatrix}
    u_1^2 & u_1 u_2 & \cdots & u_1 u_n \\
    u_2 u_1 & u_2^2 & \cdots & u_2 u_n \\
    \vdots & \vdots & \ddots & \vdots \\
    u_n u_1 & u_n u_2 & \cdots & u_n^2
    \end{bmatrix}_{n \times n}
    $$
    *Note: $uu'$ is the variance-covariance matrix of errors (unnormalized).*

### Weighted Sum of Squares
If $W_{(n \times n)}$ is diagonal, the product involving $u'_{(1 \times n)}$ and $u_{(n \times 1)}$ is the weighted sum of squares:
$$
u'Wu = (u_1, \dots, u_n) \begin{bmatrix}
w_1 & 0 & \cdots & 0 \\
0 & w_2 & \cdots & 0 \\
\vdots & \vdots & \ddots & \vdots \\
0 & 0 & \cdots & w_n
\end{bmatrix} \begin{pmatrix} u_1 \\ \vdots \\ u_n \end{pmatrix} = \sum_{i=1}^n w_i u_i^2
$$
*   **OLS** estimation minimizes the unweighted MSE $u'u$.
*   **GLS** minimizes the weighted version $u'Wu$ to improve efficiency (handling heteroskedasticity).

## 3. Regression Applications

### Identification
Typically, a matrix $X$ is not square; we need $n > k$ to identify and estimate a linear regression model.
*   The product $X'_{(k \times n)}X_{(n \times k)}$ is a **square** $k \times k$ symmetric matrix.
*   The product $X'y$ is a $k \times 1$ column vector.

### The Identity Vector
The identity vector is a column of 1s:
$$
\mathbf{i}_{(n \times 1)} = \begin{pmatrix} 1 \\ 1 \\ \vdots \\ 1 \end{pmatrix}
$$
Properties:
*   $\mathbf{i}'\mathbf{i} = n$
*   $\mathbf{i}'x_{(n \times 1)} = \sum x_i$
*   $(\mathbf{i}' \mathbf{i})^{-1}\mathbf{i}\mathbf{i}' = \begin{bmatrix} 1/n & \cdots & 1/n \\ \vdots & \ddots & \vdots \\ 1/n & \cdots & 1/n \end{bmatrix}$

### Sample Mean
The sample mean can be expressed as:
$$
(\mathbf{i}' \mathbf{i})^{-1}\mathbf{i}'x = \frac{1}{n}\sum x_i = \bar{x}
$$

### Centering (Demeaning) a Matrix
Since $x = Ix$, we can center data using the **centering matrix** $M^0$:
$$
\begin{pmatrix} x_1 - \bar{x} \\ x_2 - \bar{x} \\ \vdots \\ x_n - \bar{x} \end{pmatrix} = [x - (\mathbf{i}' \mathbf{i})^{-1} \mathbf{i} \mathbf{i}' x] = [I - (\mathbf{i}' \mathbf{i})^{-1} \mathbf{i} \mathbf{i}']x = M^0x
$$
Where $M^0$ is:
$$
M^0 = I - (\mathbf{i}' \mathbf{i})^{-1}\mathbf{i}\mathbf{i}' = \begin{bmatrix}
1-1/n & -1/n & \cdots & -1/n \\
-1/n & 1-1/n & \cdots & -1/n \\
\vdots & \vdots & \ddots & \vdots \\
-1/n & -1/n & \cdots & 1-1/n
\end{bmatrix}
$$
$M^0$ is **symmetric** and **idempotent** ($M^0 M^0 = M^0$).

#### Sum of Squared Deviations
Using $M^0$, we can express the sum of squared deviations:
$$
\sum_{i=1}^n (x_i - \bar{x})^2 = (M^0 x)'(M^0 x) = x'M^{0'}M^0x = x'M^0x
$$
Similarly for covariance:
$$
\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y}) = x'M^0y
$$

## 4. Inverse Matrices

### Definition
For a scalar $a$, the inverse is $a^{-1} = 1/a$ such that $a \times a^{-1} = 1$ (if $a \neq 0$).
For a $k \times k$ matrix $A$, the inverse $A^{-1}$ exists if and only if $A^{-1}A = I_k$ and $AA^{-1} = I_k$.

**Conditions:**
*   $A$ is **invertible** $\iff |A| \neq 0 \iff \text{rank}(A) = k$
*   $A$ is **singular** $\iff |A| = 0 \iff \text{rank}(A) < k$
*(Where $|A|$ is the determinant and $\text{rank}(A)$ is the matrix rank).*

### Solving for $\beta$ (OLS Estimator)
We solve for $\beta$ by multiplying $(X'X)^{-1}$ to both sides of the normal equation $X'y = X'X\hat{\beta}$:
$$
(X'X)^{-1} X'y = (X'X)^{-1} X'X\hat{\beta}
$$
$$
\Rightarrow (X'X)^{-1} X'y = I\hat{\beta} = \hat{\beta}
$$
Thus, the OLS estimator is $\hat{\beta} = (X'X)^{-1} X'y$.

### Properties of Inverse Matrices
1.  If a matrix has an inverse, it is **unique**.
2.  $(\alpha A)^{-1} = \frac{1}{\alpha}A^{-1}$ for scalar $\alpha \neq 0$ (if $A^{-1}$ exists).
3.  $(A^{-1})^{-1} = A$.
4.  $(A')^{-1} = (A^{-1})'$.
5.  $(AB)^{-1} = B^{-1}A^{-1}$ (where both are square and invertible).   


## Tutorial Exercise

Now that we've breifly explored the background of a regression, lets do some exercises. 

Firstly, lets load in the necessary R packages we will need for these exercises.

```{r load packages}
rm(list = ls())

library(haven) # helps us read in the janky .dta files we're using
library(dplyr) # helps manipulate data structures such as data.frames
library(data.table) # a much better and faster data.frame
```

## Question 2

### #1. load data

```{r load data 1}
data <- read_dta("~/Projects/regression_analysis/1_regression/econmath.dta")
data <- na.omit(data)
```

### #2. check the head of the table

```{r show head data 1}
head(data)
```

### #3. find the min of variable "acteng", "actmth" and "colgpa".

```{r minimums}
min(data$acteng)
min(data$actmth)
min(data$colgpa)
min(data$score)
```

### #4. find the max

```{r maximums}
max(data$acteng)
max(data$actmth)
max(data$colgpa)
max(data$score)
```

### #5. find the mean

```{r means}
mean(data$acteng)
mean(data$actmth)
mean(data$colgpa)
mean(data$score)
```

### #6. find the sd

```{r standard deviations}
sd(data$acteng)
sd(data$actmth)
sd(data$colgpa)
sd(data$score)
```

### #7. estimate the OLS model: $$score = \beta_0 + \beta_1 \cdot colgpa + \beta_2 \cdot actmth + \beta_3 \cdot acteng + u$$
```{r ols1}
OLS1 = lm(score ~ colgpa + actmth + acteng , data=data)

summary(OLS1)
```

### #8. standardize all variables and re-estimate the OLS model: 

$$score = \beta_0 + \beta_1 \cdot colgpa + \beta_2 \cdot actmth + \beta_3 \cdot acteng + u$$
```{r standardised ols 1}

data_st = select(data, score, colgpa, actmth, acteng )

data_st$score  = (data_st$score  - mean(data_st$score ))/sd(data_st$score )
data_st$colgpa = (data_st$colgpa - mean(data_st$colgpa))/sd(data_st$colgpa)
data_st$actmth = (data_st$actmth - mean(data_st$actmth))/sd(data_st$actmth)
data_st$acteng = (data_st$acteng - mean(data_st$acteng))/sd(data_st$acteng)

OLS2 = lm(score ~ colgpa + actmth + acteng , data=data_st)

summary(OLS2)
```

## Question 3

### #1. load data

```{r read in data2 }
data <- read.csv("~/Projects/regression_analysis/1_regression/hprice1.csv")
data <- na.omit(data)
```

### #2. check the head of the table

```{r head of data 2}
head(data)
```

### #3. find the min of variable "price", "sqrmtr", "lotsize" and 'bdrms'.

```{r mins}
min(data$price)
min(data$sqrft)
min(data$bdrms)
```

### #4. find the max

```{r maxs}
max(data$price)
max(data$sqrfg)
max(data$bdrms)
```

### #5. find the mean

```{r avg}
mean(data$price)
mean(data$sqrft)
mean(data$bdrms)
```

### #6. find the mean

```{r stds}
sd(data$price)
sd(data$sqrft)
sd(data$bdrms)
```

### #7. estimate the OLS model: 
$$price = \beta_0 + \beta_1 \cdot sqrmtr + \beta_2 \cdot lotsize + \beta_3 \cdot bdrms + u$$
```{r ols models 2}
OLS1 = lm(price ~ sqrft + bdrms , data=data)

summary(OLS1)
```

### #8. standardize all variables and re-estimate the OLS model
```{r standadised model 2}

data_st = select(data, price, sqrft, bdrms)

data_st$price = (data_st$price - mean(data_st$price))/sd(data_st$price)
data_st$sqrft = (data_st$sqrft - mean(data_st$sqrft))/sd(data_st$sqrft)
data_st$bdrms = (data_st$bdrms - mean(data_st$bdrms))/sd(data_st$bdrms)

OLS2 = lm(price ~ sqrft + bdrms , data=data_st)

summary(OLS2)
```

### #9. predict the value of the house
```{r predictions 1}

new_obs <- data.frame(sqrft = 2438,
                      bdrms = 4)
predict(OLS1, new_obs)

```

### #10. compute the prediction residual if price = 300
```{r predictions 2}
new_obs <- data.frame(price = 300, 
                      sqrft = 2438,  
                      bdrms = 4)

300 - predict(OLS1, new_obs)

```
