---
title: "Panel Data Introduction"
output: 
  html_document:
    toc: true
    toc_float: true
---

<style>
/* Table of Contents Styles */
#TOC {
  color: #333333;
  font-family: 'Arial', sans-serif; 
  font-size: 16px; 
  border-color: #333333;
}

/* Header Styles */
#header {
  color: #B22222; 
  background-color: #F9F9F9; 
  opacity: 0.9; 
  font-family: 'Arial', sans-serif;
  font-size: 24px;
  padding: 15px; 
}

/* Body Styles */
body {
  color: #000000; 
  font-family: 'Arial', sans-serif;
  background-color: #FFFFFF; 
  line-height: 1.6; 
}

/*  Text Styles */
pre {
  color: #4D4D4D; 
  background-color: #F0F8FF; 
  border: 1px solid #B0C4DE; 
  padding: 10px; 
}

/* Header Styles */
h1, h2, h3 {
  color: #003366;
  font-family: 'Arial', sans-serif;
}

/* Link Styles */
a {
  color: #003366; 
  text-decoration: none; 
}

a:hover {
  text-decoration: underline; 
}
</style>










```{r load packages, message=FALSE}
rm(list = ls()) 

library(dplyr)
library(data.table) 
library(ggplot2)
library(fastDummies)
library(plm)

options(scipen = 999) # makes numbers look nicer
```

Great now lets load in some data for us to play around with. 

This dataset is a very simple collection of house sales. We will try to model some prices to see which features of houses are most important in determining prices *(as well as how helpful they are at predicting house prices)*.

```{r data load}

df <- setDT(read.csv("https://raw.githubusercontent.com/Michael-Morgan-Giles/regression_analysis/refs/heads/main/data/Co2.csv"))

df[, gdp_per_cap := gdp / population]
df[, ln_gdp_per_cap := log(gdp_per_cap)]
df[, ln_co2_per_capita := log(co2_per_capita)]
df[, ln_co2 := log(co2)]
df[, ln_feed_in_tariff := log(Feed...in...Tariff + 1)]
df[, CP := Effective.Carbon.Price..Sum.of..carbon.price.rate...proportion.of.countries.CO2.covered..]
df[, ln_CP := log(CP + 1)]
df[, CP_DV := Carbon_price_DV]

# set up panel
panel <- pdata.frame(df, index = c('iso_code', 'year'))

```



```{r formula}
formula <- as.formula("ln_co2_per_capita ~ CP_DV + ln_gdp_per_cap + ln_feed_in_tariff + ln_CP")
```


```{r pooled ols}
pols <- plm(formula, 
            data = panel, 
            model = "pooling")

summary(pols, vcov = vcovHC(pols, type = "HC1", cluster = "group"))

```



```{r random effects}
re_mod <- plm(formula, 
            data = panel, 
            model = "random",
            effect = "individual")

summary(re_mod, vcov = vcovHC(re_mod, type = "HC1", cluster = "group"))

```



```{r fixed effects}
fe_mod <- plm(formula, 
            data = panel, 
            model = "within",
            effect = "twoway")

summary(fe_mod, vcov = vcovHC(fe_mod, type = "HC1", cluster = "group"))

```



```{r hausman test}
phtest(fe_mod, re_mod)

```





```{r coefficient table, warning=FALSE}
stargazer(pols, fe_mod, re_mod,
          title="Model Comparisons",
          header=FALSE, 
          type="text",  
          omit.table.layout="n",
          digits=3, 
          intercept.bottom=TRUE, 
          dep.var.labels.include = FALSE,
          model.numbers = FALSE,
          dep.var.caption="Dependent variable: ln_co2_per_capita",
          model.names=FALSE,
          #omit.stat=c("LL","ser"),
          column.labels=c("Pooled OLS", "FE", "RE"))

```





