---
title: "Difference-in-differences Introduction"
output: 
  html_document:
    toc: true
    toc_float: true
---

<style>
/* Table of Contents Styles */
#TOC {
  color: #333333;
  font-family: 'Arial', sans-serif; 
  font-size: 16px; 
  border-color: #333333;
}

/* Header Styles */
#header {
  color: #B22222; 
  background-color: #F9F9F9; 
  opacity: 0.9; 
  font-family: 'Arial', sans-serif;
  font-size: 24px;
  padding: 15px; 
}

/* Body Styles */
body {
  color: #000000; 
  font-family: 'Arial', sans-serif;
  background-color: #FFFFFF; 
  line-height: 1.6; 
}

/*  Text Styles */
pre {
  color: #4D4D4D; 
  background-color: #F0F8FF; 
  border: 1px solid #B0C4DE; 
  padding: 10px; 
}

/* Header Styles */
h1, h2, h3 {
  color: #003366;
  font-family: 'Arial', sans-serif;
}

/* Link Styles */
a {
  color: #003366; 
  text-decoration: none; 
}

a:hover {
  text-decoration: underline; 
}
</style>










```{r load packages, message=FALSE}
rm(list = ls()) 

library(dplyr)
library(data.table) 
library(ggplot2)
library(fastDummies)
library(plm)

options(scipen = 999) # makes numbers look nicer
```



```{r data load}

df <- setDT(read.csv("https://raw.githubusercontent.com/Michael-Morgan-Giles/python-causality-handbook/refs/heads/master/causal-inference-for-the-brave-and-true/data/smoking.csv"))

df[, treated := fifelse(california == "True", 1, 0)]
df[, post_treat := fifelse(after_treatment == "True", 1, 0)]
df[, treated_post_treat := treated * post_treat]



```


```{r plot grouped}
df[california == "False"] %>% 
  ggplot(aes(x = year, y = cigsale, color = as.factor(state)))+
  geom_line()+
  geom_line(data = df[california == "True"], aes(y = cigsale), color = "red")+
  theme_classic()+
  theme(legend.position = "blank")+
  labs(title = "Historical Cigarette Sales Per Capita by State")+
  geom_vline(xintercept = 1988, color = "black", type = "--")+
  scale_color_viridis_d()

```


```{r plot even more grouped, message=FALSE}
df[, .(mean_cigsale = mean(cigsale)), by = .(year, california)] %>% 
  ggplot(aes(x = year, y = mean_cigsale, color = california))+
  geom_line()+
  theme_classic()+
  theme(legend.position = "blank")+
  labs(title = "Historical Cigarette Sales Per Capita by State")+
  geom_vline(xintercept = 1988, color = "black", type = "--")+
  geom_smooth(method = "lm", data = df[year < 1989, .(mean_cigsale = mean(cigsale)), by = .(year, california)], se = FALSE)

```


```{r manual calcs}

mu_after_treat = df[treated == 1 & post_treat == 1, mean(cigsale)]
mu_before_treat = df[treated == 1 & post_treat == 0, mean(cigsale)]
mu_after_control = df[treated == 0 & post_treat == 1, mean(cigsale)]
mu_before_control = df[treated == 0 & post_treat == 0, mean(cigsale)]

did_manual <- (mu_after_treat - mu_before_treat) - (mu_after_control - mu_before_control)

print(paste0(' The manual DiD estimator is: ', round(did_manual,2)))

```


```{r pooled ols}
did_ols <- lm(cigsale ~ treated + post_treat + treated_post_treat, 
            data = df)

summary(did_ols, vcov = vcovHC(did_ols, cluster = "group"))

```






```{r fixed effects calcs}
fe_mod <- plm(cigsale ~ treated_post_treat, 
            data = df, 
            model = "within",
            effect = "twoway",
            index = c("state", "year"))

summary(fe_mod, vcov = vcovHC(fe_mod, type = "HC1", cluster = "group"))

```



```{r heterogenous}

df <- dummy_cols(df, select_columns = "year")

formula <- as.formula(paste0("cigsale ~ ", 
                             paste0("treated:post_treat:year_",
                                    as.character(levels(as.factor(df$year))),
                                    sep = "", 
                                    collapse = "+")))
  
dyn_did <- plm(formula, 
            data = df, 
            model = "within",
            effect = "twoway",
            index = c("state", "year"))

summary(dyn_did, vcov = vcovHC(dyn_did, cluster = "group"))
```





```{r dynamic plot, warning=FALSE}

cbind(year = df[year > 1987, year], summary(dyn_did, 
                              vcov = vcovHC(dyn_did, 
                                            cluster = "group"))$coefficients) %>%
  ggplot(aes(x = year, y = Estimate))+
  geom_line(color = "black")+
  geom_ribbon(aes(ymin = Estimate - 2 * `Std. Error`, 
                  ymax = Estimate + 2 * `Std. Error`), 
              fill = "cornflowerblue", 
              color = "black", 
              alpha = 0.5)+
  geom_hline(yintercept = 2, linetype = "dashed", color = "black")+
  theme_classic()+
  labs(title = "Heterogenous DiD Estimates",
       y = "DiD Estimate")+
  theme(legend.position = "top")

```





