---
title: "Treatment Effects with Panel Data"
output: 
  html_document:
    toc: true
    toc_float: true
---

<style>
/* Table of Contents Styles */
#TOC {
  color: #333333;
  font-family: 'Arial', sans-serif; 
  font-size: 16px; 
  border-color: #333333;
}

/* Header Styles */
#header {
  color: #B22222; 
  background-color: #F9F9F9; 
  opacity: 0.9; 
  font-family: 'Arial', sans-serif;
  font-size: 24px;
  padding: 15px; 
}

/* Body Styles */
body {
  color: #000000; 
  font-family: 'Arial', sans-serif;
  background-color: #FFFFFF; 
  line-height: 1.6; 
}

/*  Text Styles */
pre {
  color: #4D4D4D; 
  background-color: #F0F8FF; 
  border: 1px solid #B0C4DE; 
  padding: 10px; 
}

/* Header Styles */
h1, h2, h3 {
  color: #003366;
  font-family: 'Arial', sans-serif;
}

/* Link Styles */
a {
  color: #003366; 
  text-decoration: none; 
}

a:hover {
  text-decoration: underline; 
}
</style>

## Difference-in-differences

Now that we've gone through the introduction to panel data esitmators in the last notebook, it's time to look at a very common estimator - **Difference-in-differences (DiD)**. 

DiD is a quasi-experimental method that is often employed to estimate a causal effect when randomisation is not possible. It requires a panel / longitudinal dataset that follows multiple units *(at least two)* over time. 

At a high level, DiD compares the outcomes of one group that recieves some treatment *(say an increase in the excise tax on tabacco to pick one random example)*, to another group that does not recieve that treatment. I will refer to these groups as the treatment and control groups respectively. 

A bit more specifically, the DiD estimator compares the mean of each groups outcome both before and after the treatment is applied *(hence the name, difference-in-differences)*. Now that seems like quite a simple approach *(and it is)*, however it requires some key assumptions. 

The most essential of these assumptions is the known as the "parallel trends" assumption. This requires that there are no "time-varying" characteristics that are different between the two groups that could explain the differences in the outcomes. For example, one group might experience a recession in the economy whilst the other didn't. 

This is the essential assumption. If there are some unobserved effect that impact the groups across time, the estimated effect cannot be entirely attributed to the treatment itself *(it's possible, and likely, that time-varying difference explain the change in outcome instead)*. 

Now, that is probably a little confusing at this stage. So the best way to learn is to look at an example! 

Let's return to the example of the California Excise tax on Tabacoo that we explored in the time series notebook. 

To refresh quickly in 1989, proposition 99 was enacted that increased the excise tax on cigarettes by 25 cents per pack. In theory, this increase in cost should decrease the number of cigarettes consumed. That is a falsifiable statement. 

Now, really handily this excise tax was only enacted in California, yet all the other states in the US did not enact this excise tax. These states are a potential pool of control "units".

So, now that we know a bit more about panel data, we can compare the time series of cigarette sales per capita before and after intervention in both California *(where the treatment occoured i.e. treatment group)* and those states where there was no increase in the excise on tabaco *(the control group)*.

With background out of the way, lets show how we can calculate the DiD estimator.

```{r load packages, message=FALSE}
rm(list = ls()) 

library(dplyr)
library(data.table) 
library(ggplot2)
library(fastDummies)
library(plm)
library(lmtest)

options(scipen = 999) # makes numbers look nicer
```

Below we load in the data and create some variables that will be handy when calculating the DiD estimator.

```{r data load}

df <- setDT(read.csv("https://raw.githubusercontent.com/Michael-Morgan-Giles/python-causality-handbook/refs/heads/master/causal-inference-for-the-brave-and-true/data/smoking.csv"))

df[, treated := fifelse(california == "True", 1, 0)]
df[, post_treat := fifelse(after_treatment == "True", 1, 0)]
df[, treated_post_treat := treated * post_treat]

```

# How do we calculate the DiD Estimator?
## Intervention Timeline

The first step in good DiD implementation is to visually check the parallel trends assumption. There are a few ways to do this, however the simplest is to compare each units time series for the outcome variable, `cigsales` in our example representing cigarette sales per capita.

If we plot each of the states cigarette sales per capita over the available years and compare this too California's time series *(in red)*, we can see that there is quite a bit of difference across the states. 

In fact, California already has one of the lowest cigarette consumption per capita rates of all the states. 

This might raise a problem. How can we be certain that our estimates are because the excise was introduced, and not because cigarette sales were already on a downward trend for some other *(potentially unobserved)* reason? 

```{r plot grouped}
df[california == "False"] %>% 
  ggplot(aes(x = year, y = cigsale, color = as.factor(state)))+
  geom_line()+
  geom_line(data = df[california == "True"], aes(y = cigsale), color = "red")+
  theme_classic()+
  theme(legend.position = "blank")+
  labs(title = "Historical Cigarette Sales Per Capita by State")+
  geom_vline(xintercept = 1988, color = "black", type = "--")+
  scale_color_viridis_d()

```

Before we answer that question, we can simplify the above plot to see if, on average, the trends between the two groups are different. If we plot the mean of each groups we can see that there does appear to be a different trend between the two groups. It also appears to be quite substantial. 

Now, this means the DiD estimator is probably biased. However, we can still calculate the estimate, noting that this is likely an upper bound of the estimate.

A good question for learning might be how to ask how we know that this will be an upper bound estimate *(hint: think about where the two trends prior to intervention will end up, and what this might mean for the final estimate)*.

```{r plot even more grouped, message=FALSE}
df[, .(mean_cigsale = mean(cigsale)), by = .(year, california)] %>% 
  ggplot(aes(x = year, y = mean_cigsale, color = california))+
  geom_line()+
  theme_classic()+
  theme(legend.position = "blank")+
  labs(title = "Historical Cigarette Sales Per Capita by State")+
  geom_vline(xintercept = 1988, color = "black", type = "--")+
  geom_smooth(method = "lm", data = df[year < 1989, .(mean_cigsale = mean(cigsale)), by = .(year, california)], se = FALSE)

```

## Manual calculation

Okay, so understanding that there is some bias, lets try get a point estimate anyhow *(using DiD)*. As mentioned in the introduction, the easiest way to estimate DiD is by comparing the group means pre and post treatment. 

We can set the notation fo the mean as $ \mu_{i,t}$ where $i \in \text{treatment, control}$ and where $t \in \text{before treatment, after treatment}$. 

The DiD estimate is then the difference in the treatment means before and after, minus the difference in the control means before and after i.e:

$$ (\mu_{\text{treated,after}} - \mu_{\text{treated,before}}) - (\mu_{\text{control,after}} - \mu_{\text{control,before}})$$

This is quite easy to code up:

```{r manual calcs}

mu_after_treat = df[treated == 1 & post_treat == 1, mean(cigsale)]
mu_before_treat = df[treated == 1 & post_treat == 0, mean(cigsale)]
mu_after_control = df[treated == 0 & post_treat == 1, mean(cigsale)]
mu_before_control = df[treated == 0 & post_treat == 0, mean(cigsale)]

did_manual <- (mu_after_treat - mu_before_treat) - (mu_after_control - mu_before_control)

print(paste0(' The manual DiD estimator is: ', round(did_manual,2)))

```

To get a better understanding of how this calculation is working, see if you can get this same estimate from the table below: 

```{r manual calcs table}

dcast(df[, mean(cigsale), by = .(treated, post_treat)],treated ~ post_treat, value.var = "V1")

```

### What is happening under the hood?

Okay, the above is all nice and well, but what is actually happening with this estimator? Well at a high level, the DiD estimator is infering a "counterfactual". A counterfactual is what would have happened to the treated unit in the abscence of the treatment i.e. no excise tax introduced in California. 

We don't know what the counterfactual is because we only observe the "actual" i.e. California did have an increase in the excise tax. 

So we have to estimate this counterfactual. This is what the DiD estimator is actually doing. A visualisation might be illustrative here:

```{r counterfactual viz}

df[, .(mean_cigsale = mean(cigsale)), 
   by = .(treated, post_treat)] %>% 
  ggplot(aes(x = post_treat, y = mean_cigsale, color = as.factor(treated)))+
  geom_line()+
  geom_point(size = 3)+
  geom_line(data = data.frame(post_treat = c(0,1),
                              cf = c(mu_before_treat, mu_before_treat + (mu_after_control - mu_before_control))),
            aes(x = post_treat, y = cf), 
            inherit.aes = FALSE,
            colour = "lightblue",
            linetype = "dashed",
            show.legend = TRUE)+   
  labs(title = "DiD Estimator") +
  theme_classic()+
  theme(legend.position = "none")

```

The graph above show the means of each group *(treated = solid blue, control = salmon)* for the before (0) and after (1) treatment periods. This actually represents the mean points in the table above. 

The new dashed blue line shows the inferred counterfactual the DiD estimator calculates. The DiD effect is the difference between the end of the solid blue line, and the end of the blue dotted line i.e: 

$$ \text{Causal Effect = Actual - Counterfactual}$$

You might notice that the counterfactual in this scenario basically looks like the control group, but shifted down to start from the treated groups initial mean value. You would be right to think that is interesting. A good follow up question might be how does the parallel trend assumption apply here? *(hint: is it reasonable to assume that this counterfactual will follow the same trend as the control group? What did we see about the trends of each group prior to intervention above?)*.

## OLS
Now that we've explored what's going on under the hood, lets look at some other helpful ways to calculate the DiD estiamtor. Perhaps the most common way is to use OLS. Very handily, the above calcualtion can be simplified into a linear model. 

It involves the defining the binary variables:

$$
\ AfterTreatment_t = \begin{cases}
    1 & \text{if } t \text{ is after treatment} \\
    0 & \text{otherwise.}
\end{cases}
$$

and:

$$
\ Treated_i = \begin{cases}
    1 & \text{if } i \text{ is the treated unit} \\
    0 & \text{otherwise.}
\end{cases}
$$

If we put these variables into a matrix as such: 

$$
X = \begin{bmatrix}
1 & Teated_1 & AfterTreatment_1 & Treated_1 \times AfterTreatment_1 \\
1 & Teated_2 & AfterTreatment_2 & Treated_2 \times AfterTreatment_2 \\
1 & Teated_3 & AfterTreatment_3 & Treated_3 \times AfterTreatment_3 \\
\vdots & \vdots & \vdots & \vdots  \\
1 & Teated_i & AfterTreatment_t & Treated_i \times AfterTreatment_t
\end{bmatrix}
$$

We can use the OLS estimator to calculate $\beta'$, and $\beta_3$ of that vector will be the DiD estimate.

Let's code this up:

```{r pooled ols}
did_ols <- lm(cigsale ~ treated + post_treat + treated_post_treat, 
            data = df)

coeftest(did_ols, vcov. = sandwich::vcovCL(did_ols, cluster = ~state))

```

As you can see the coefficeint for `treated_post_treat` is the same as the manual calculations above. If you look at the table below, you can see how each scenario corresponds to the four possible scenarios from the manual calculations: 

| Scenario | Time Period | Excise Tax Increase | Conditions ($Treated_i, AfterTreatment_t$) | Coefficient |
 | :--- | :--- | :--- | :--- | :--- | 
 | Control Group (Pre) | pre 1988 | No | $$Treated_i=0, AfterTreatment_t=0$$ | $$\beta_0$$ |
 | Control Group (Post) | post 1988 | No | $$Treated_i=0, AfterTreatment_t=1$$ | $$\beta_0 + \beta_2$$ | 
| Treatment Group (Pre) |  pre 1988 | Yes | $$Treated_i=1, AfterTreatment_t=0$$ | $$\beta_0 + \beta_1$$ |
| Treatment Group (Post) | post 1988 | Yes | $Treated_i=1, AfterTreatment_t=1$ | $\beta_0 + \beta_1 + \beta_2 + \beta_3$ | 


If we take the above and apply each set of coefficients to the manual calculations we get:

$$
DID = (\mu_{\text{treated,after}} - \mu_{\text{treated,before}}) - (\mu_{\text{control,after}} - \mu_{\text{control,before}})
$$

$$
DiD = [(\beta_0 + \beta_1 + \beta_2 + \beta_3) - (\beta_0 + \beta_1)] - [(\beta_0 + \beta_2) - \beta_0]
$$

$$
DiD = \beta_2 + \beta_3 - \beta_2 = \beta_3
$$

## Fixed Effects (TWFE)


Note in the above that we get estimates for all of the the elements in $\beta$. This is obviously fine, but say we wanted to just get the $\beta_3$ coefficient?

Well, if we apply the "within" transformation to remove the "time- & unit-invariant" coefficients *(and effects)*, to calculate just the DiD estimate. Instead of including everything in the $X$ matrix above, we just include the varibale `treated_post_treat` and apply the within transformation to estimate $\beta_3$. 

Let's code this up!

```{r fixed effects calcs}
fe_mod <- plm(cigsale ~ treated_post_treat, 
            data = df, 
            model = "within",
            effect = "twoway",
            index = c("state", "year"))

coeftest(fe_mod, vcov. = vcovHC(fe_mod, type = "HC1",cluster = "group"))

```

Yay, it's the same!

It's worth noting that none of these above estimates are "better" than one another. They are estimating the same effect and hence will have the same results *(note all three methods produce the 27.34 decrease in cigarette sales per capita)*. The differences / benefits of each approach depends on what estimates you want to produce. By way of example, lets take a look at a popular choice using the TWFE approach that can estimate "dynamic" or "Heterogenous" effects.

# Heterogenous DiD Estimates

We can calculate dynamic / heterogenous DiD estimates we can fit the following model via TWFE:

$$ Y_{it} = \sum_{k\neq-1} \beta_k \times 1\{Treated = 1\} \times 1\{t = k\} + \alpha_i + \lambda_t + \varepsilon_{it}$$

Where $k$ is the treatment / intervention timeline i.e. $k=0$ is the time the intervention occours. In the above model we also have the typical time and unit fixed effects, yet we also have this interaction term where our estimates for $\beta_k$ are interaction with a dummary varibale for the treated unit, $1\{Treated = 1\}$, and dummy variables for each of the time periods in $k$, $1\{t = k\}$, i.e. a 1 for each time period in $k$, otherwise a zero. 

Note that $\beta_{r=-1}$ is not estimated because it acts as a "baseline year". It is really helpful to use TWFE in this model because all the other time and unit invariant terms will cancel out when we estimate using TWFE. 

This means our model will produce one coefficient for each period in $k$, excluding the baseline year. This can make investigating how the effect develops over time quite simple *(and visual)*, hence why it has become quite a popular approach. So, let's try and code this up!

```{r heterogenous}


df <- dummy_cols(df, select_columns = "year")

formula <- as.formula(paste0("cigsale ~ ", 
                             paste0("treated:post_treat:year_",
                                    as.character(levels(as.factor(df$year))),
                                    sep = "", 
                                    collapse = "+")))

# if that formula looks confusing, print the below code to see what it's doing
#print(formula)
dyn_did <- plm(formula, 
            data = df, 
            model = "within",
            effect = "twoway",
            index = c("state", "year"))

summary(dyn_did, vcov = vcovHC(dyn_did, cluster = "group"))
```

Amazing! we can see that the immediate effect was around 10 cigarette sales per capita less, followed by a steady decline through to the year 2000.

This has become quite a popular estimator in econometrics, especially because it is quite easy to plot these estimates!

```{r dynamic plot, warning=FALSE}

cbind(year = df[year > 1987, year], summary(dyn_did, 
                              vcov = vcovHC(dyn_did, 
                                            cluster = "group"))$coefficients) %>%
  ggplot(aes(x = year, y = Estimate))+
  geom_line(color = "black")+
  geom_ribbon(aes(ymin = Estimate - 2 * `Std. Error`, 
                  ymax = Estimate + 2 * `Std. Error`), 
              fill = "cornflowerblue", 
              color = "black", 
              alpha = 0.5)+
  geom_hline(yintercept = 2, linetype = "dashed", color = "black")+
  theme_classic()+
  labs(title = "Heterogenous DiD Estimates",
       y = "DiD Estimate")+
  theme(legend.position = "top")

```
