---
title: "Endogeneity and Instrumental Variables"
output: 
  html_document:
    toc: true
    toc_float: true
---

<style>
/* Table of Contents Styles */
#TOC {
  color: #333333;
  font-family: 'Arial', sans-serif; 
  font-size: 16px; 
  border-color: #333333;
}

/* Header Styles */
#header {
  color: #B22222; 
  background-color: #F9F9F9; 
  opacity: 0.9; 
  font-family: 'Arial', sans-serif;
  font-size: 24px;
  padding: 15px; 
}

/* Body Styles */
body {
  color: #000000; 
  font-family: 'Arial', sans-serif;
  background-color: #FFFFFF; 
  line-height: 1.6; 
}

/*  Text Styles */
pre {
  color: #4D4D4D; 
  background-color: #F0F8FF; 
  border: 1px solid #B0C4DE; 
  padding: 10px; 
}

/* Header Styles */
h1, h2, h3 {
  color: #003366;
  font-family: 'Arial', sans-serif;
}

/* Link Styles */
a {
  color: #003366; 
  text-decoration: none; 
}

a:hover {
  text-decoration: underline; 
}
</style>

## What is Endogeneity?







```{r load packages, message=FALSE}
rm(list = ls()) 

library(dplyr)
library(haven) # helps us read .dta files
library(data.table) 
library(stargazer)
library(AER)
library(ggplot2)

options(scipen = 999) # makes numbers look nicer
```

Great now lets load in some data for us to play around with. 

This dataset is a very simple collection of house sales. We will try to model some prices to see which features of houses are most important in determining prices *(as well as how helpful they are at predicting house prices)*.

```{r data load}

df <- setDT(read_dta("https://raw.githubusercontent.com/Michael-Morgan-Giles/regression_analysis/refs/heads/main/data//maketable4.dta"))
## only keep base sample - only reproducing some esitmates
df <- df[baseco == 1]

df[, oth_cont := fifelse(shortnam %in% c('AUS', 'MLT', 'NZL'), 1, 0 )]
```

Below is a table describing the data. 
| Name | Description |
| :--- | :--- |
| `shortnam` | 3-letter country name |
| `africa` | Africa |
| `latitude` | Absolute value of latitude |
| `rich4` | Country is in neo-Europe  |
| `avexpr` | Exp Risk |
| `logpgp95` |  Log GDP per capita (PPP) in 1995  |
| `logem4` |  Original Log Settler Mortality  |
| `asia` | Asia |
| `loghjypl` |  Log output per worker in 1988  |
| `baseco` | =1 if part of base sample |
| `oth_cont`| =1 if other continent | 

Note that this data was downloaded from a MIT dropbox on 16/6/2026 -https://www.dropbox.com/scl/fi/3yuv9j514zuajzjfluoc1/maketable4.zip?rlkey=pq9l7bxktw1iqxe6fmoh26g79&e=1&dl=0


```{r table 2 column 2, message=FALSE}
t2c2_mod <- lm(logpgp95 ~ avexpr, data = df)

summary(t2c2_mod)

df %>% 
  ggplot(aes(x = avexpr, y = logpgp95)) +
  geom_point()+
  geom_smooth(method = 'lm')+
  theme_classic()
```





```{r table 4 Panel B column 1, message=FALSE}

t4pbc1_mod <- lm(avexpr ~ logem4, data = df)

summary(t4pbc1_mod)

df %>% 
  ggplot(aes(x = logem4, y = avexpr)) +
  geom_point()+
  geom_smooth(method = 'lm')+
  theme_classic()
```



```{r Table 4 Panel A column 1 }
t4pAc1_IV_mod <- ivreg(logpgp95 ~ avexpr | logem4, data = df)

summary(t4pAc1_IV_mod)

```


```{r Table 4 Panel A column 7}
t4pAc7_IV_mod <- ivreg(logpgp95 ~ avexpr + asia + africa + oth_cont | 
                         logem4 + asia + africa + oth_cont,
                       data = df)

summary(t4pAc7_IV_mod)
```

```{r table 4 Panel B column 8}
t4pAc8_IV_mod <- ivreg(logpgp95 ~ avexpr + lat_abst + asia + africa + oth_cont | 
                         logem4 + lat_abst +asia + africa + oth_cont,
                       data = df)

summary(t4pAc8_IV_mod)
```

```{r stargrazer tests, warning=FALSE}
stargazer(t2c2_mod, t4pbc1_mod, t4pAc1_IV_mod, t4pAc7_IV_mod, t4pAc8_IV_mod,
          title = "Model Comparison",
          header = FALSE,
          type = "text",
          omit.table.layout = "n",
          digits = 4,
          intercept.bottom = TRUE,
          column.labels = c('Table 2 Col 2',
                            'Table 4 Panel B Col 1',
                            'Table 4 Panel A Col 1',
                            'Table 4 Panel A Col 7',
                            'Table 4 Panel A Col 8'),
          dep.var.caption = "Log GDP per Capita 1995",
          dep.var.labels.include = FALSE,
          model.numbers = TRUE,
          model.names = FALSE
          )
```



